%! Author = Len Washington III
%! Date = 9/8/25

% Preamble
\documentclass[
	lecture={01},
	title={Overview}
]{enve201notes}

% Packages

% Document
\begin{document}

\maketitle

%<*Lecture-1>
\chapter{Overview}\label{ch:overview}

\section{Study of Earth}\label{sec:study-of-earth}
\begin{itemize}
	\item \textit{Geology} (also called \textit{geoscience}) is the study of the Earth.
	\item It encompasses studies that characterize:
	\begin{itemize}
		\item The formation and composition of this planet.
		\item The causes of mountain building and ice ages.
		\item The record of life's evolution.
		\item The history of climate change
		\item Practical problems such as pollution, resources recovery, and natural disaster
	\end{itemize}
	\item Understanding natural Earth processes better prepare us to thrive in our \emph{environment}.
\end{itemize}

\section{Environment}\label{sec:environment}
\begin{itemize}
	\item The word ``Environment'' is derived from the French word ``Environ'' which means ``surrounding''.
	\item Environment is a collection of systems including all living and nonliving things on Earth that sustain all life -- \textit{complex of many variables}!
	\item Environment includes water, air, land, human beings and other living creatures such as plants, animals and micro-organisms.
	\item The natural environment consist of five interlinking systems namely, the geosphere, the hydrosphere, the cryosphere, the biosphere, and the atmosphere.
\end{itemize}

\section{Earth System at a Glance}\label{sec:earth-system-at-a-glance}
\begin{itemize}
	\item The \textbf{geosphere} (or lithosphere) consists of the solid part of our planet.
	\item The \textbf{hydrosphere} consists of all water at or near the surface of the Earth.
	\item The \textbf{cryosphere} consists of frozen water, mostly in glaciers.
	\item The \textbf{biosphere} consists of living organisms, from bacteria to whales.
	\item The \textbf{atmosphere} is the envelope of gas that encircles the planet.
\end{itemize}

\emph{Earth system interacts!}
%</Lecture-1>

\end{document}